---
title: "Cleaning-Bad-Data-in-R"
autor: "Luis Fernández"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
	message = FALSE,
	warning = FALSE)
```

## 1.2 Missing Fields

After loading the heating data I need to transform and format some values. When I apply mutate to transform homes `heating %>% mutate(homes = as.numeric(homes))` from Class: character, R show me a \textcolor{red}{Caused by warning: ! NAs introducidos por coerción}. Then I ask for the data using a filter and found characters values (".", "Z"). Finally mutate the values to zero after check with other sources. \

```{r library, echo = TRUE,	message = FALSE,	warning = FALSE}
# Load the tidyverse
library(tidyverse)
```



```{r exc1, include=TRUE, echo=TRUE, warning = TRUE}

# Load the data file
heating <- read_csv("./exercise_files/1_2/heating.csv")

# Tidy the data
heating <- heating %>% gather(key="age", value="homes", -Source)

knitr::kable(
  summary(heating),
  digits=1, align=rep('c', 5))
  
knitr::kable(
  head(heating %>% mutate(homes = as.numeric(homes)), 10),
   digits=1, align=rep('c', 5))

knitr::kable(
  heating %>% filter(is.na(as.numeric(homes))),
  digits=1, align=rep('c', 5))

```


```{r exc2, echo=TRUE, warning = TRUE}

heating <- 
  heating %>%
  mutate(homes=ifelse(homes=='.', 0, homes)) %>% 
  mutate(homes=ifelse(homes=='Z', 0, homes)) %>% 
  mutate(homes = as.numeric(homes))

  summary(heating)

```


