---
title: "Cleaning-Bad-Data-in-R"
autor: "Luis Fernández"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
	message = FALSE,
	warning = FALSE)
```

# 1. Missing Data

## 1.2. Missing Fields

After loading the heating data I need to transform and format some values. When I apply mutate to transform homes `heating %>% mutate(homes = as.numeric(homes))` from Class: character, R show me a \textcolor{red}{Caused by warning: ! NAs introducidos por coerción}. Then I ask for the data using a filter and found characters values (".", "Z"). Finally mutate the values to zero after check with other sources. 
\

```{r library, echo = TRUE,	message = FALSE,	warning = FALSE}
# Load the tidyverse
library(tidyverse)
```


```{r exc1, include=TRUE, echo=TRUE, warning = TRUE}

# Load the data file
heating <- read_csv("./exercise_files/1_2/heating.csv")

# Tidy the data
heating <- heating %>% gather(key="age", value="homes", -Source)

knitr::kable(
  summary(heating),
  digits=1, align=rep('c', 5))
  
knitr::kable(
  head(heating %>% mutate(homes = as.numeric(homes)), 7),
   digits=1, align=rep('c', 5))

knitr::kable(
  heating %>% filter(is.na(as.numeric(homes))),
  digits=1, align=rep('c', 5))

```


```{r exc2, echo=TRUE, warning = TRUE}

heating <- 
  heating %>%
  mutate(homes=ifelse(homes=='.', 0, homes)) %>% 
  mutate(homes=ifelse(homes=='Z', 0, homes)) %>% 
  mutate(homes = as.numeric(homes))

  summary(heating)

```

\newpage

## 1.3. Missing Rows

\colorbox{yellow}{In this exercise I have lands acres from USA, but there are just 42 obs, however the count of states are 50.} \
\colorbox{yellow}{So is necessary create a data table in tiddy format with the other ones.} \

```{r Missing Rows, echo=TRUE, warning = TRUE}

# Load the data file
land <- read_csv("./exercise_files/1_3/publiclands.csv")

summary(land)
nrow(land)
unique(land$State)

missing_states <- tibble(State= c('Connecticut', 'Delawere', 'Hawai', 'Iowa', 'Maryland', 
                                  'Massachusetts', 'New Jersey', 'Rhode Island'),
                         PublicLandAcres=c(0,0,0,0,0,0,0,0))

land <- rbind(land, missing_states)

knitr::kable(
  tail(land, 10),
    digits=1, align=rep('c', 5))

```

\newpage

## 1.4. Agregation and Missing Values

\colorbox{yellow}{Sometimes when we calculate some statistics like sum or count, the result is NA because} \
\colorbox{yellow}{R do not have all the values. In this case we need to add na.rm = TRUE.}\

```{r Agregation, echo=TRUE, warning = TRUE}

# Load the data file
employees <- read_csv("./exercise_files/1_4/employees.csv")

knitr::kable(employees, digits=1, align=rep('c', 5))

sum(employees$Salary)
mean(employees$Salary)
max(employees$Salary)

sum(employees$Salary, na.rm = TRUE)
mean(employees$Salary, na.rm = TRUE)
max(employees$Salary, na.rm = TRUE)

```
\newpage

